\documentclass[11pt,letterpaper]{article}

\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{ebnf2tikz}
\ebnffile{manual-examples.ebnf}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  urlcolor=blue!70!black,
  pdftitle={ebnf2tikz User Manual},
  pdfauthor={Larry D. Pyeatt}
}

\lstdefinelanguage{ebnf}{
  morekeywords={},
  morestring=[b]',
  morestring=[b]",
  morecomment=[l]{//},
  morecomment=[l]{---},
  morecomment=[s]{(*}{*)},
  literate={|}{{\textbar}}1
           {<<--}{{\textless\textless--}}4
           {-->>}{{--\textgreater\textgreater}}4,
  sensitive=true
}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  stringstyle=\color{red!70!black},
  commentstyle=\color{green!50!black}\itshape,
  showstringspaces=false,
  frame=single,
  framerule=0.4pt,
  xleftmargin=1em,
  xrightmargin=1em,
  breaklines=true,
  aboveskip=1em,
  belowskip=1em
}

\newcommand{\opt}[1]{\texttt{#1}}
\newcommand{\prog}{\texttt{ebnf2tikz}}

\title{\textbf{ebnf2tikz User Manual}}
\author{Larry D. Pyeatt}
\date{\today}

\begin{document}

\maketitle
\tableofcontents
\newpage

%% ----------------------------------------------------------------
\section{Introduction}

\prog{} is an optimizing compiler that converts Extended Backus-Naur
Form (EBNF) grammar specifications into railroad (syntax) diagrams
expressed as \LaTeX{} TikZ commands.  It supports annotations that
control inlining of productions, figure captions, and rotated figures.
An optimizer simplifies the resulting abstract syntax tree before
layout, producing compact, readable diagrams.

\prog{} is free software distributed under the GNU General Public
License, version~3 or later.

%% ----------------------------------------------------------------
\section{Installation}

\subsection{Prerequisites}

\begin{itemize}
  \item A C++ compiler with C++17 support (e.g., \texttt{g++})
  \item \href{https://www.gnu.org/software/bison/}{Bison} $\geq$ 3.7
  \item \href{https://github.com/westes/flex}{Flex}
  \item \href{https://cmake.org/}{CMake} $\geq$ 3.16
  \item \texttt{pdflatex} with TikZ (for rendering diagrams)
\end{itemize}

\subsection{Building}

\begin{lstlisting}[language=bash]
cmake .
make
\end{lstlisting}

This produces the \prog{} executable, \texttt{ebnf2tikz.sty} (the
\LaTeX{} package), and \texttt{ebnf2tikz.pdf} (package documentation)
in the project root.  Additional targets:

\begin{lstlisting}[language=bash]
make check        # Run AST-dump regression tests
make check-tikz   # Build PDF with all railroad diagrams
make check-sty    # Run LaTeX package integration tests
make sty          # Extract ebnf2tikz.sty from .dtx
make sty-doc      # Build ebnf2tikz.pdf package documentation
make doc          # Generate Doxygen API documentation
make realclean    # Remove all build artifacts
\end{lstlisting}

%% ----------------------------------------------------------------
\section{Quick Start}

\begin{enumerate}
\item Create a \LaTeX{} document that loads the package and includes
  an inline grammar:
\begin{lstlisting}[language={[LaTeX]TeX}]
\documentclass{article}
\usepackage{ebnf2tikz}
\begin{document}

\begin{ebnf}
greeting = 'hello' , [ name ] , '!' ;
name = identifier ;
\end{ebnf}

The greeting syntax is:
\useproduction{greeting}

\end{document}
\end{lstlisting}

\item Compile three times with shell escape enabled:
\begin{lstlisting}[language=bash]
pdflatex -shell-escape mydocument
pdflatex -shell-escape mydocument
pdflatex -shell-escape mydocument
\end{lstlisting}
\end{enumerate}

This multi-pass workflow is similar to the feedback loop required by
Bib\TeX{} or \texttt{makeindex}.  On the first pass, the \texttt{ebnf}
environment writes the grammar to a file and \prog{} runs with
estimated node dimensions.  \LaTeX{} then measures the actual rendered
sizes of each node and writes them to \texttt{bnfnodes.dat}.  On
subsequent passes, \prog{} reads these measurements and produces an
accurate layout.  See Section~\ref{sec:latex} for details.

%% ----------------------------------------------------------------
\section{Command-Line Options}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Short & Long & Description \\
\midrule
\opt{-O} & \opt{-{}-optimize}     & Enable optimization (default). \\
\opt{-n} & \opt{-{}-nooptimize}   & Disable optimization and subsumption. \\
\opt{-f} & \opt{-{}-makefigures}  & Wrap \texttt{tikzpicture}s in \LaTeX{} \texttt{figure} environments. \\
\opt{-d} & \opt{-{}-dumponly}     & Parse and dump the AST; do not generate TikZ output. \\
\opt{-B} & \opt{-{}-dump-before}  & Dump the AST before optimization (to stdout). \\
\opt{-A} & \opt{-{}-dump-after}   & Dump the AST after optimization (to stdout). \\
\opt{-h} & \opt{-{}-help}         & Print usage information and exit. \\
\bottomrule
\end{tabular}
\end{center}

\medskip\noindent
Usage:
\begin{lstlisting}[language=bash]
./ebnf2tikz [options] input_file output_file
\end{lstlisting}

The \opt{-O} flag is the default and is accepted for compatibility.
Use \opt{-n} to suppress both the optimizer and annotation-driven
subsumption.

The \opt{-B} and \opt{-A} flags print the AST to standard output
before and after optimization, respectively.  These are useful for
debugging grammars.  The \opt{-d} flag prints the AST and skips TikZ
generation entirely.

%% ----------------------------------------------------------------
\section{EBNF Syntax Reference}

\prog{} accepts a practical subset of the EBNF notation defined by
ISO/IEC~14977~\cite{iso14977}, with several extensions for railroad
diagram generation.  See Appendix~\ref{sec:iso14977} for a detailed
comparison with the standard.

\subsection{Productions}

A grammar consists of one or more productions.  Each production has a
name, an equals sign, a body expression, and a terminating semicolon:

\begin{lstlisting}[language=ebnf]
production_name = body_expression ;
\end{lstlisting}

\subsection{Constructs}

\begin{center}
\begin{tabular}{@{}llp{6cm}@{}}
\toprule
Syntax & Meaning & Example \\
\midrule
\texttt{A , B}     & Sequence (concatenation) & \texttt{'if' , expr , 'then'} \\
\texttt{A | B}     & Choice (alternation)     & \texttt{'true' | 'false'} \\
\texttt{[ A ]}     & Optional                 & \texttt{[ 'else' , statement ]} \\
\texttt{\{ A \}}   & Repetition (zero or more) & \texttt{\{ ',' , param \}} \\
\texttt{( A )}     & Grouping                 & \texttt{( 'a' | 'b' ) , 'c'} \\
\texttt{'text'}    & Terminal (single quotes)  & \texttt{'while'} \\
\texttt{"text"}    & Terminal (double quotes)  & \texttt{"begin"} \\
\texttt{? text ?}  & Special sequence          & \texttt{? any character ?} \\
\texttt{name}      & Nonterminal reference     & \texttt{expression} \\
\texttt{\textbackslash\textbackslash} & Explicit line break & Forces a new row in the diagram \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Comments}

Three comment styles are supported:

\begin{lstlisting}[language=ebnf]
// C++-style line comment
--- Ada-style line comment
(* Block comment, can span
   multiple lines *)
\end{lstlisting}

%% ----------------------------------------------------------------
\section{Annotations}
\label{sec:annotations}

Annotations are placed immediately before a production definition
using the \verb|<<-- ... -->>| syntax.  Multiple annotations may be
placed before a single production.

\subsection{\texttt{subsume}}

\begin{lstlisting}[language=ebnf]
<<-- subsume -->>
production_name = body ;
\end{lstlisting}

The \texttt{subsume} annotation marks a production for inlining:
wherever its name appears as a nonterminal reference in other
productions, the reference is replaced by the production's body.  The
subsumed production itself is not emitted as a separate diagram.

\subsection{\texttt{subsume as}}

\begin{lstlisting}[language=ebnf]
<<-- subsume as regex_pattern -->>
production_name = body ;
\end{lstlisting}

Like \texttt{subsume}, but the production is only inlined into
nonterminals whose names match the given POSIX extended regular
expression.

\subsection{\texttt{caption}}

\begin{lstlisting}[language=ebnf]
<<-- caption "A descriptive caption" -->>
production_name = body ;
\end{lstlisting}

Sets the \LaTeX{} figure caption for this production's diagram.
Requires the \opt{-f} flag to have any effect, since captions only
apply to \texttt{figure} environments.

\subsection{\texttt{label}}

\begin{lstlisting}[language=ebnf]
<<-- label "fig:my-production" -->>
production_name = body ;
\end{lstlisting}

Sets the \LaTeX{} \verb|\label| for this production's figure, allowing
cross-references with \verb|\ref{fig:my-production}|.  Requires the
\opt{-f} flag.  The \verb|\label| is always emitted after \verb|\caption|
in the generated output.  Multiple annotations can be combined with
semicolons:

\begin{lstlisting}[language=ebnf]
<<-- caption "The greeting syntax" ; label "fig:greeting" -->>
greeting = 'hello' , [ name ] , '!' ;
\end{lstlisting}

\subsection{\texttt{sideways}}

\begin{lstlisting}[language=ebnf]
<<-- sideways -->>
production_name = body ;
\end{lstlisting}

Places this production's diagram in a \texttt{sidewaysfigure}
environment (from the \texttt{rotating} package) instead of a regular
\texttt{figure}.  Requires the \opt{-f} flag.  Can be combined with
\texttt{caption} and \texttt{label}:

\begin{lstlisting}[language=ebnf]
<<-- sideways ; caption "Full SQL syntax" ; label "fig:sql" -->>
select_stmt = 'SELECT' , column_list , 'FROM' , table_ref ;
\end{lstlisting}

%% ----------------------------------------------------------------
\section{\LaTeX{} Integration}
\label{sec:latex}

The \texttt{ebnf2tikz} \LaTeX{} package provides all required TikZ
styles, commands, lengths, and the multi-pass infrastructure.  Install
\texttt{ebnf2tikz.sty} where \LaTeX{} can find it (e.g., in your
document directory or your local \texttt{texmf} tree), then add:

\begin{lstlisting}[language={[LaTeX]TeX}]
\usepackage{ebnf2tikz}
\end{lstlisting}

\subsection{Package Options}

\begin{center}
\begin{tabular}{@{}lp{9cm}@{}}
\toprule
Option & Effect \\
\midrule
\texttt{figures}    & Pass \opt{-f} to \prog{}, wrapping each diagram in a \texttt{figure} environment. \\
\texttt{nooptimize} & Pass \opt{-n} to \prog{}, disabling optimization and subsumption. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Providing EBNF Input}

There are two ways to provide EBNF grammars to the package.

\paragraph{External files.}
Register one or more \texttt{.ebnf} files in the preamble or body:

\begin{lstlisting}[language={[LaTeX]TeX}]
\ebnffile{mygrammar.ebnf}
\end{lstlisting}

\paragraph{Inline grammars.}
Use the \texttt{ebnf} environment to write grammars directly in your
\LaTeX{} source:

\begin{lstlisting}[language={[LaTeX]TeX}]
\begin{ebnf}
greeting = 'hello' , [ name ] , '!' ;
name = identifier ;
\end{ebnf}
\end{lstlisting}

Content inside the \texttt{ebnf} environment is captured verbatim and
written to \verb|\jobname-rail.ebnf|.

\subsection{Placing Diagrams}

Use \verb|\useproduction{name}| to place a specific production's
railroad diagram:

\begin{lstlisting}[language={[LaTeX]TeX}]
The greeting syntax is:
\useproduction{greeting}
\end{lstlisting}

The name must match a production name from your EBNF input.  When
using \opt{-f} (or the \texttt{figures} package option), diagrams are
automatically wrapped in \texttt{figure} environments and do not
require \verb|\useproduction|.

\subsection{Multi-Pass Workflow}

With \texttt{-shell-escape} enabled, the package runs \prog{}
automatically at \verb|\begin{document}|.  The typical workflow is:

\begin{lstlisting}[language=bash]
pdflatex -shell-escape mydocument   # captures EBNF, runs ebnf2tikz
pdflatex -shell-escape mydocument   # uses real node sizes from pass 1
pdflatex -shell-escape mydocument   # final accurate layout
\end{lstlisting}

Without \texttt{-shell-escape}, run \prog{} manually between passes:

\begin{lstlisting}[language=bash]
pdflatex mydocument
ebnf2tikz mydocument-rail.ebnf mydocument-rail.tex
pdflatex mydocument
ebnf2tikz mydocument-rail.ebnf mydocument-rail.tex
pdflatex mydocument
\end{lstlisting}

\subsection{Customizable Lengths}

The package defines several lengths with sensible defaults.  Override
them with \verb|\setlength| after loading the package:

\begin{center}
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
Length & Default & Purpose \\
\midrule
\verb|\railcolsep|     & 5pt            & Horizontal gap between nodes. \\
\verb|\railrowsep|     & \verb|\baselineskip| & Vertical gap between rows in multi-row diagrams. \\
\verb|\railnodeheight| & 14pt           & Minimum height of terminal and nonterminal nodes. \\
\verb|\railcorners|    & 4pt            & Rounding radius for rail corners. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Customizable Styles}
\label{sec:styles}

The package defines three TikZ styles that control the visual
appearance of railroad diagrams.

\paragraph{The \texttt{raildiagram} style.}
Every railroad diagram is wrapped in
\verb|\begin{tikzpicture}[raildiagram]|.  The \texttt{raildiagram}
style sets the default line weight to \texttt{thick} for all rail
connections, and establishes a baseline node style with zero padding
and no visible border:

\begin{lstlisting}[language={[LaTeX]TeX}]
raildiagram/.style={
  thick,
  every node/.style={inner sep=0, outer sep=0, draw=none}
}
\end{lstlisting}

This style is \emph{scoped} to railroad diagram pictures only.  It
does not affect any other \texttt{tikzpicture} environments in your
document.  You can freely mix railroad diagrams with circuit diagrams,
plots, commutative diagrams, or any other TikZ content without style
interference.

\paragraph{The \texttt{terminal} style.}
Draws a rounded rectangle with a subtle grey gradient fill.  Applied
to terminal (literal) tokens such as \texttt{'while'} or
\texttt{"begin"}.

\paragraph{The \texttt{nonterminal} style.}
Draws a sharp-cornered rectangle with a red-tinted gradient fill.
Applied to nonterminal references such as \texttt{expression} or
\texttt{statement}.

\paragraph{Font commands.}
Three commands control the fonts used for text within diagrams:

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Command & Default & Purpose \\
\midrule
\verb|\railtermname|   & \verb|\bfseries| & Terminal node text \\
\verb|\railname|       & \verb|\itshape|  & Nonterminal node text \\
\verb|\railprodname|   & \verb|\itshape|  & Production name labels \\
\bottomrule
\end{tabular}
\end{center}

Use the setter commands \verb|\setrailtermfont|,
\verb|\setrailnontermfont|, and \verb|\setraillabelfont| to change
these after loading the package.

%% ----------------------------------------------------------------
\section{Auto-Wrap}
\label{sec:autowrap}

Productions whose body exceeds the available \verb|\textwidth| are
automatically broken across multiple rows in the railroad diagram.
This happens without any user intervention.

Additionally, wide nonterminal labels that contain spaces are
automatically wrapped using \verb|\shortstack| to reduce their
rendered width.

If you need to force a line break at a specific point, use the
explicit line break operator \verb|\\| in your EBNF input:

\begin{lstlisting}[language=ebnf]
long_production =
    first_part , second_part , \\ third_part , fourth_part ;
\end{lstlisting}

%% ----------------------------------------------------------------
\section{Examples}
\label{sec:examples}

The examples in this section are live---they are rendered by
\prog{} and \LaTeX{} as part of building this manual.

%% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Simple Productions}

A basic grammar with a sequence, a choice, and an optional element:

\begin{lstlisting}[language=ebnf]
// Simple productions
manual_greeting = 'hello' , [ manual_recipient ] , '!' ;
manual_recipient = 'world' | manual_name ;
manual_name = first_name , [ last_name ] ;
\end{lstlisting}

\noindent Rendered output:

\medskip\noindent
\useproduction{manual_greeting}

\medskip\noindent
\useproduction{manual_recipient}

\medskip\noindent
\useproduction{manual_name}

%% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Repetition (Loops)}

Repetition is expressed with curly braces.  The optimizer transforms
loops with separators into a compact form:

\begin{lstlisting}[language=ebnf]
// Repetition
manual_number = manual_digit , { manual_digit } ;
manual_parameter_list = parameter , { ',' , parameter } ;
manual_digit = '0'|'1'|'2'|'3'|'4'|'5'|'6'|'7'|'8'|'9' ;
\end{lstlisting}

\noindent Rendered output:

\medskip\noindent
\useproduction{manual_number}

\medskip\noindent
\useproduction{manual_parameter_list}

\medskip\noindent
\useproduction{manual_digit}

%% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Subsumption}

The \texttt{subsume} annotation inlines a production wherever it is
referenced, then the optimizer simplifies the result.  Consider this
VHDL fragment:

\begin{lstlisting}[language=ebnf]
case_statement_alternative =
    'when' , choices , '=>', sequence_of_statements;

<<-- subsume -->>
choices = choice, { '|', choice } ;

<<-- subsume -->>
choice = simple_expression | discrete_range
       | simple_name | 'others' ;
\end{lstlisting}

\noindent
After subsumption and optimization, only a single compact diagram is
produced for \texttt{manual\_case\_alt}:

\medskip\noindent
\useproduction{manual_case_alt}

\medskip\noindent
Without subsumption, the diagram would instead show
\texttt{choices} and \texttt{choice} as opaque nonterminal
references, requiring the reader to look up separate diagrams.

%% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Complex Example with Nested Constructs}

This example combines sequences, choices, optionals, and loops:

\begin{lstlisting}[language=ebnf]
select_stmt = 'SELECT' , [ 'DISTINCT' ] ,
    column_list , 'FROM' , table_ref ,
    [ 'WHERE' , condition ] ,
    [ 'ORDER' , 'BY' , order_list ] ;
\end{lstlisting}

\noindent Rendered output:

\medskip\noindent
\useproduction{manual_select_stmt}

%% ----------------------------------------------------------------
\section{Customization Examples}
\label{sec:customization}

All railroad diagrams are generated inside
\verb|\begin{tikzpicture}[raildiagram]|, so the \texttt{raildiagram}
TikZ style controls the overall appearance of every diagram.  Because
it is applied as an option to individual \texttt{tikzpicture}
environments, it is completely scoped---changes to \texttt{raildiagram}
affect only railroad diagrams, never other TikZ pictures in the
document.

\subsection{Default Appearance}

With no customization, the default \texttt{raildiagram} style uses
\texttt{thick} lines, terminals are grey rounded rectangles, and
nonterminals are pink-tinted rectangles:

\begin{lstlisting}[language={[LaTeX]TeX}]
% No customization needed --- this is the default
\usepackage{ebnf2tikz}
\end{lstlisting}

All of the diagrams in Section~\ref{sec:examples} use the default
appearance.

\subsection{Changing Line Weight}

To make railroad diagrams use thinner lines, append to the
\texttt{raildiagram} style:

\begin{lstlisting}[language={[LaTeX]TeX}]
\tikzset{raildiagram/.append style={thin}}
\end{lstlisting}

To change the line weight for individual node types as well, also
update the \texttt{terminal} and \texttt{nonterminal} styles:

\begin{lstlisting}[language={[LaTeX]TeX}]
\tikzset{
  raildiagram/.append style={thin},
  terminal/.append style={thin},
  nonterminal/.append style={thin}
}
\end{lstlisting}

\subsection{Changing Node Colors}

Override the gradient fills on terminal and nonterminal nodes:

\begin{lstlisting}[language={[LaTeX]TeX}]
\tikzset{
  terminal/.append style={
    top color=white, bottom color=blue!20
  },
  nonterminal/.append style={
    top color=white, bottom color=green!20
  }
}
\end{lstlisting}

For flat (non-gradient) fills, replace the shading keys entirely:

\begin{lstlisting}[language={[LaTeX]TeX}]
\tikzset{
  terminal/.style={
    draw, inner sep=2pt, rounded rectangle,
    minimum size=\railnodeheight, thick,
    fill=orange!15
  },
  nonterminal/.style={
    draw, inner sep=2pt, rectangle,
    minimum size=\railnodeheight, thick,
    fill=cyan!15
  }
}
\end{lstlisting}

\subsection{Changing Fonts}

Use the setter commands to change the font used for text inside nodes
and production labels:

\begin{lstlisting}[language={[LaTeX]TeX}]
\setrailtermfont{\ttfamily\bfseries}     % monospace bold terminals
\setrailnontermfont{\sffamily\itshape}   % sans-serif italic nonterminals
\setraillabelfont{\sffamily\bfseries}    % sans-serif bold labels
\end{lstlisting}

\subsection{Changing Spacing}

Adjust the spacing parameters after loading the package:

\begin{lstlisting}[language={[LaTeX]TeX}]
\setlength{\railcolsep}{8pt}            % wider horizontal gaps
\setlength{\railrowsep}{20pt}           % taller vertical gaps
\setlength{\railnodeheight}{18pt}       % taller nodes
\setlength{\railcorners}{6pt}           % rounder corners
\end{lstlisting}

Note: after changing these lengths, you must re-run the full
\prog{} $\to$ \texttt{pdflatex} feedback loop, since the layout
engine reads spacing parameters from the node sizes file.

\subsection{Replacing the raildiagram Style Entirely}

If you need full control, you can completely replace the
\texttt{raildiagram} style.  When doing so, you \emph{must} include
the baseline \texttt{every node} reset, because the generated TikZ
code depends on nodes having zero default padding:

\begin{lstlisting}[language={[LaTeX]TeX}]
\tikzset{
  raildiagram/.style={
    semithick,
    every node/.style={inner sep=0, outer sep=0, draw=none}
  }
}
\end{lstlisting}

Omitting the \texttt{every node} reset will cause unwanted padding
and borders on the coordinate nodes and connection points that form
the railroad tracks.

\subsection{Coexistence with Other TikZ Packages}

Because \texttt{raildiagram} is scoped to railroad pictures, it does
not interfere with other TikZ-based packages.  For example, you can
use \texttt{ebnf2tikz} alongside \texttt{circuitikz},
\texttt{pgfplots}, or \texttt{tikz-cd} in the same document without
any style conflicts:

\begin{lstlisting}[language={[LaTeX]TeX}]
\documentclass{article}
\usepackage{ebnf2tikz}
\usepackage{circuitikz}     % uses its own TikZ styles
\usepackage{pgfplots}       % uses its own TikZ styles
\ebnffile{mygrammar.ebnf}

\begin{document}

Railroad diagram (uses raildiagram style):
\useproduction{my_rule}

Circuit diagram (unaffected by ebnf2tikz):
\begin{circuitikz}
  \draw (0,0) to[R, l=$R_1$] (2,0) to[C, l=$C_1$] (2,-2) -- (0,-2);
\end{circuitikz}

Plot (unaffected by ebnf2tikz):
\begin{tikzpicture}
  \begin{axis}[width=6cm, height=4cm]
    \addplot {x^2};
  \end{axis}
\end{tikzpicture}

\end{document}
\end{lstlisting}

%% ----------------------------------------------------------------
\section{Tips}
\label{sec:tips}

\begin{itemize}
\item Use the \opt{-B} and \opt{-A} flags to inspect the AST before
  and after optimization.  This is helpful for understanding how the
  optimizer transforms your grammar.

\item When adding \prog{} diagrams to an existing document, it is
  convenient to \verb|\input| the generated TikZ file rather than
  pasting it inline.

\item The \opt{-f} flag wraps each diagram in a \texttt{figure}
  environment, which enables captions, labels, and
  \verb|\listoffigures|.  Without \opt{-f}, diagrams are placed
  inline.

\item Production names must be valid identifiers (letters, digits,
  underscores).  Underscores in production names are rendered in the
  diagram using \LaTeX{} formatting.

\item The feedback loop (\prog{} $\to$ \texttt{pdflatex} $\to$
  \prog{} $\to$ \texttt{pdflatex}) need only be repeated when
  node content changes.  If you are only editing prose around
  existing diagrams, a single \texttt{pdflatex} run suffices.
\end{itemize}

%% ----------------------------------------------------------------
\appendix
\section{Conformance with ISO/IEC 14977}
\label{sec:iso14977}

ISO/IEC~14977:1996~\cite{iso14977} defines the international standard
for Extended Backus-Naur Form.  \prog{} implements a practical subset
of that standard and adds several extensions for railroad diagram
generation.  This appendix details the differences.

\subsection{Supported ISO~14977 Features}

The following features conform to the standard:

\begin{center}
\begin{tabular}{@{}llp{6.5cm}@{}}
\toprule
Feature & Syntax & Notes \\
\midrule
Defining symbol       & \texttt{=}               & \\
Terminator            & \texttt{;}               & \\
Concatenation         & \texttt{,}               & \\
Definition separator  & \texttt{|}                & \\
Grouping              & \texttt{( \ldots{} )}    & \\
Optional              & \texttt{[ \ldots{} ]}    & \\
Repetition            & \texttt{\{ \ldots{} \}}  & Zero or more repetitions. \\
First quote           & \texttt{'\ldots{}'}      & Terminal strings. \\
Second quote          & \texttt{"\ldots{}"}      & Terminal strings. \\
Special sequence      & \texttt{? \ldots{} ?}    & Rendered as a terminal node. \\
Block comments        & \texttt{(* \ldots{} *)}  & May span multiple lines. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Unsupported ISO~14977 Features}

The following features defined by the standard are \emph{not} supported:

\begin{center}
\begin{tabular}{@{}llp{6.5cm}@{}}
\toprule
Feature & ISO~14977 Syntax & Notes \\
\midrule
Exception         & \texttt{A - B}     & Set subtraction of syntactic terms. \\
Repetition factor & \texttt{3 * A}     & Integer prefix for fixed repetition count. \\
Alternative brackets & \texttt{(/ /)} for \texttt{[ ]},
                       \texttt{(: :)} for \texttt{\{ \}} & Compatibility
                       notations for keyboards lacking square and curly
                       brackets.  Obsolete in practice. \\
Alternative defining symbol & \texttt{:} & Some dialects use a colon
                       instead of \texttt{=}. \\
Alternative terminator & \texttt{.}    & Some dialects use a period
                       instead of \texttt{;}. \\
Alternative separators & \texttt{/} or \texttt{!} & Some dialects use
                       these instead of \texttt{|}. \\
Spaces in meta-identifiers & \texttt{case statement} & ISO~14977
                       allows spaces in nonterminal names.  \prog{}
                       requires underscores (e.g.,
                       \texttt{case\_statement}), which are rendered as
                       spaces in the diagram. \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The \emph{exception} and \emph{repetition factor} operators are the
most significant omissions.  The exception operator (\texttt{A - B})
specifies that the syntax defined by~\texttt{A} excludes anything
matching~\texttt{B}.  This concept has no natural visual
representation in a railroad diagram, so it is omitted.  The
repetition factor (\texttt{3 * A}) specifies exactly~$n$ repetitions
of a syntactic term; railroad diagrams typically represent such
patterns as loops or sequences, making a dedicated operator
unnecessary.

The alternative bracket notations (\texttt{(/ /)}, \texttt{(: :)})
and alternative operator symbols (\texttt{:}, \texttt{.}, \texttt{/},
\texttt{!}) were included in the standard for compatibility with
character sets that lack certain punctuation.  They are obsolete on
modern systems and are not supported.

\subsection{\prog{} Extensions Beyond ISO~14977}

\prog{} extends standard EBNF with the following features:

\begin{center}
\begin{tabular}{@{}lp{8.5cm}@{}}
\toprule
Feature & Description \\
\midrule
Explicit line break (\texttt{\textbackslash\textbackslash})
  & Forces a new row in the railroad diagram at a specific point in a
    production body. \\
Annotations (\texttt{<<-{}- \ldots{} -{}->>})
  & Per-production metadata controlling subsumption, figure captions,
    labels, and rotation.  See Section~\ref{sec:annotations}. \\
Line comments (\texttt{//} and \texttt{-{}-{}-})
  & C++ and Ada style line comments, in addition to the standard
    \texttt{(* \ldots{} *)} block comments. \\
Underscore identifiers
  & Nonterminal names use underscores instead of spaces
    (e.g., \texttt{case\_statement} rather than \texttt{case statement}).
    Underscores are rendered as spaces in the diagram. \\
\bottomrule
\end{tabular}
\end{center}

\begin{thebibliography}{1}
\bibitem{iso14977}
  ISO/IEC~14977:1996,
  \emph{Information technology---Syntactic metalanguage---Extended BNF},
  International Organization for Standardization, 1996.
\end{thebibliography}

\end{document}
